A Spectral Analysis of the Domain Decomposed Monte Carlo Method for Linear Systems

### 2015-04-14 doc/ned2015/response_to_reviewer.txt

Response to reviewer's comments:

------------------------------------------------------------------------------
1. Since the underlying analysis seems to be graphical, they should use as
many graphs to explain what is happening.  Following words and moving
information is not so easy for a reader.

We agree that without visual aids the discussion is complex and is very much
based on the underlying graph of the matrix. We have added a figure (Figure 1)
in section 3.1 and discussion the random walk in terms of the sparse graph
underneath the matrices that should help clarify the explanation, particularly
for the n_s and n_i parameters in our diffusion analysis. In addition, we have
expanded our description of the algorithm in section 2 and have provided more
definitions for clarity.

To improve clarity of the results, we also have replotted our results using a
different plotting tool to create higher quality images that should aid the
reader. Note that the y-axis in figure 2 has been changed to a logarithmic
scale and the random walk length and error measurements have been broken into
two different plots.

### 2015-04-14 doc/ned2015/slattery_wilson_evans.tex
finishing ned edits

  \begin{keyword}
    Monte Carlo \sep domain decomposition \sep MCSA \sep parallel \sep
    computing \sep linear solvers
  \end{keyword}

@@

An alternative approach to solving linear systems is to employ Monte Carlo
methods that sample a distribution with an expectation value equivalent to
that of the inverted linear operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that, for certain classes of problems, are competitive with
conventional methods methods and are applicable to a wide variety of linear
systems \cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA
methods to engineering problems, our research aims to improve the underlying
Monte Carlo algorithms to handle large-scale problems in parallel computing
environments.

@@

In the Neumann-Ulam method, a Monte Carlo game is played to construct the
terms of the Neumann series in Eq~(\ref{eq:adjoint_neumann_solution}). We use
random walks of length $k$ to build the series of state transitions in the
summation:
\begin{equation}
  \nu = i_0 \rightarrow i_1 \rightarrow i_2 \rightarrow \cdots \rightarrow
  i_{k-1}\:,
  \label{eq:permutation}
\end{equation}
where $\nu$ is a particular random walk permutation. Starting a random walk
permutation requires the computation of an initial starting state, $i_0$, and
initial weight, $W_0$. The initial state, $i_0$, of the random walk is
determined by sampling the source vector $\ve{b}$ with probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
and with a corresponding starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_{i_0}}{|b_{i_0}|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
After a random walk is started, we construct the remaining terms in
Eq~(\ref{eq:adjoint_neumann_solution}) using a row-scaled transition probability
matrix and corresponding transition weights:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{|h_{ji}|}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
where each row of the probability matrix, $\ve{P}_i$, gives a probability
distribution function for moving from row $i$ to a new row, $j$. As a random
walk is computed by sampling the probability matrix, every stochastic
transition from a state $i$ to a state $j$ has a corresponding weight,
$w_{ij}$, that is used to update the current weight of the random walk. At the
$m^{th}$ step of a random walk, its weight is:
\begin{equation}
  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
with $W_0$ given by Eq~(\ref{eq:adjoint_starting_weight}). To stop a random
walk, we utilize a relative weight cutoff parameter:
\begin{equation}
  W_f = W_c W_0:\,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff}. The random walk will be terminated
after $m$ steps if $W_m < W_f$.

Using our result from Eq~(\ref{eq:inner_product_constraint}) the Monte Carlo
estimator for the solution in state $i$ for a particular random walk
permutation of length $k$ is:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in the
current state, $i_m$. Finally, the expectation value using all random walk
permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded with Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_source_definition}) directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:.
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}

@@

For our numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

@@

Compared to the serial construction of the probabilities and weights, data for
all states that are required to be on-process must be collected in parallel so
that all possible random walks that leave the local domain may be correctly
communicated. As an example of walking through the sparse graph of the matrix
and parallel communication, Figure~\ref{fig:diffusion_graph} gives the
adjacency graph of a single state in the neutron diffusion matrix for the
model problem presented in Eq~(\ref{eq:fd_system}). The discretization stencil
of the diffusion problem dictates the structure of this graph and the states
to which a random walk may transition. Consider a random walk starting at node
$(i,j)$ in the mesh (the center node). If we play the Monte Carlo game to move
to a new state, then we may move to any of the nodes in this graph, including
the node at which we started with a probability proportional to the
coefficients of each node in Eq~(\ref{eq:fd_system}). If any grid point
adjacent to node $(i,j)$ is on another processor, the random walk must be
communicated to that processor to continue the walk.

@@

We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$
on a rectilinear Cartesian grid:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi preconditioning
is sufficient to reduce the spectral radius of the iteration matrix below
unity and therefore ensure convergence of the Neumann series. The
preconditioner in this case is $\ve{M} = diag(\ve{D})$ such that we are
solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion operator
with each row scaled by the diagonal component. As we have defined a
homogeneous domain, the scaling factor, $\alpha$, is the same for all rows in
the operator and defined as the $\phi_{i,j}$ coefficient from
Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we have the following spectrum of preconditioned
eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

@@

The adjoint Monte Carlo method is effectively an approximation to a
stationary method. Stationary methods for linear systems arise from
splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration
  index}. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the solution
error at the $k^{th}$ iterate, the error after $k$ iterations is:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we have:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}

@@

In a domain decomposed situation, not all random walks will remain within the
domain they started in and must instead be communicated. This communication,
expected to be expensive, was analyzed by Siegel and colleagues for idealized,
load balanced situations for full nuclear reactor core Monte Carlo simulations
\cite{siegel_analysis_2012}.  To quantify the number of particles that leak
out of the local domain they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{\text{average \# of particles leaving local domain}}
          {\text{total \# of particles starting in local domain}}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, it was assumed that the value of $\Lambda$ was dependent on
the total cross section of the system via the Wigner rational
approximation. Outlined more thoroughly by Hwang's chapter in
\cite{azmy_nuclear_2010}, we will use both the Wigner rational approximation
and the mean chord approximation as a means to estimate the leakage fraction.

In the case of parallel linear systems, we can use diffusion theory to
estimate the optical thickness of a domain in the decomposition and the
corresponding leakage fraction in terms of properties of the linear operator
and the discretization. We first calculate the mean distance a random walk
will move in the grid by computing the mean squared distance of its movement
along the chord of length $l$ defined along a cardinal direction of the
domain. After a single transition a random walk will have moved a mean squared
distance of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord and $n_s$
is the number of grid elements a random walk will move on average every
transition in the direction of the chord. For our diffusion model problem,
$n_s$ would equate to the expected number of states in the $i$ (or $j$ as the
problem is symmetric) direction that a random walk will move in a single
transition and is dependent on the stencil given in
Figure~\ref{fig:diffusion_graph}. After $k$ transitions, the random walk will
have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our chord is of length $l$ and there are $n_i$ grid elements (or
states to which a random walk may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimensionality of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo random walk in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of a domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we get:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

For optically thin domains, we expect that most random walks will be
communicated, while optically thick domains will leak the fraction of random
walks that did not interact within. Using the optical thickness defined in
Eq~(\ref{eq:optical_thickness_3}), we can complete the leakage approximations
by defining the bounds of $\tau \rightarrow 0, \Lambda \rightarrow 1$ and
$\tau \rightarrow \infty, \Lambda \rightarrow \tau^{-1}$.  With these bounds
we define the leakage fraction out of a domain using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight random walks.

@@

In future work, these relationships will serve as guidelines for
selecting an appropriate parallel algorithm strategy and provide a
basis for performance models for future parallel implementations of
the adjoint Neumann-Ulam method. To be applicable to a broader range
of problems, these models could potentially be extended to
non-symmetric systems. In addition, these relations will be used to
analyze parallel implementations of Monte Carlo synthetic acceleration
methods that leverage the adjoint method and serve as an initial
grounds for assessing their feasibility for large-scale problems.

>>>>

  \begin{keyword}
    Monte Carlo \sep domain decomposition \sep MCSA \sep parallel computing
    \sep linear solvers
  \end{keyword}

@@

An alternative approach to solving linear systems is to employ Monte Carlo
methods that sample a distribution with an expectation value equivalent to
that of the inverted linear operator. Such methods have been in existence for
decades with the earliest reference noted here an enjoyable manuscript
published in 1950 by Forsythe and Leibler \cite{forsythe_matrix_1950}. In
their outline, Forsythe and Leibler in fact credit the creation of this
technique to J. Von Neumann and S.M. Ulam some years earlier than its
publication. In 1952 Wasow provided a more formal explanation of Von Neumann
and Ulam's method \cite{wasow_note_1952} and Hammersley and Handscomb's 1964
monograph \cite{hammersley_monte_1964} and Spanier and Gelbard's 1969 book
\cite{spanier_monte_1969} present additional detail on a method adjoint to
that described by Neumann and Ulam. To improve their convergence, Halton
leveraged the Monte Carlo algorithm in an iterative refinement scheme where a
residual Monte Carlo solution serves as a correction to a fixed point
iteration \cite{halton_sequential_1962}. In recent years, the adjoint
Neumann-Ulam method has been leveraged in Monte Carlo Synthetic Acceleration
(MCSA) methods that, for certain classes of problems, are competitive with
conventional methods methods and are applicable to a variety of linear systems
\cite{evans_monte_2009,evans_monte_2014}. To apply these new MCSA methods to
engineering problems, our research aims to improve the underlying Monte Carlo
algorithms to handle large-scale problems in parallel computing environments.

@@

In the Neumann-Ulam method, a Monte Carlo game is played to construct the
terms of the Neumann series in Eq~(\ref{eq:adjoint_neumann_solution}). We use
random walks of length $k$ to build the series of state transitions in the
summation:
\begin{equation}
  \nu = i_0 \rightarrow i_1 \rightarrow i_2 \rightarrow \cdots \rightarrow
  i_{k-1}\:,
  \label{eq:permutation}
\end{equation}
where $\nu$ is a particular random walk permutation. Starting a random walk
permutation requires the computation of an initial starting state, $i_0$, and
initial weight, $W_0$. The initial state is determined by sampling the source
vector $\ve{b}$ with probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
and with a corresponding starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_{i_0}}{|b_{i_0}|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
After a random walk is started, we construct the remaining terms in
Eq~(\ref{eq:adjoint_neumann_solution}) using a row-scaled transition probability
matrix and corresponding transition weights:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,\ w_{ij} =
  \frac{|h_{ji}|}{p_{ij}},
  \label{eq:adjoint_probability}
\end{equation}
where each row of the probability matrix, $\ve{P}_i$, gives a probability
distribution function for moving from row $i$ to a new row, $j$. As a random
walk is computed by sampling the probability matrix, every stochastic
transition from a state $i$ to a state $j$ has a corresponding weight,
$w_{ij}$, that is used to update the current weight of the random walk. At the
$m^{th}$ step of a random walk, its weight is:
\begin{equation}
  W_{m} = W_0 w_{i_0,i_1} w_{i_1,i_2} \cdots w_{i_{m-1},i_m}\:,
  \label{eq:direct_permutation_weight}
\end{equation}
with $W_0$ given by Eq~(\ref{eq:adjoint_starting_weight}). To stop a random
walk, we utilize a relative weight cutoff parameter:
\begin{equation}
  W_f = W_c W_0:\,
  \label{eq:relative_weight_cutoff}
\end{equation}
where $W_c$ is the \textit{weight cutoff}. The random walk will be terminated
after $m$ steps if $W_m < W_f$.

Using our result from Eq~(\ref{eq:inner_product_constraint}) the Monte Carlo
estimator for the solution from a particular random walk permutation of length
$k$ is:
\begin{equation}
  X_{\nu} = \sum_{m=0}^k W_{m} \delta_{i,i_m}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contribution for each step
of the random walk is only in its current state, $i_m$. Finally, the
expectation value using all random walk permutations is:
\begin{equation}
  E\{X\} = \sum_{\nu} P_{\nu} X_{\nu}\:
  \label{eq:adjoint_expectation_value}
\end{equation}
which if expanded with Eqs~(\ref{eq:adjoint_probability}) and
(\ref{eq:adjoint_source_definition}) directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:.
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}

@@

For our derivations and numerical experiments, we choose the one-speed,
two-dimensional neutron diffusion equation as a model problem
\cite{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

@@

Compared to the serial construction of the probabilities and weights, domain
decomposition requires that data for all states be on-process so that all
possible random walks that leave the local domain may be correctly
communicated. As an example of walking through the sparse graph of the matrix
and parallel communication, Figure~\ref{fig:diffusion_graph} gives the
adjacency graph of a single state in the neutron diffusion matrix for the
model problem presented in Eq~(\ref{eq:fd_system}). The discretization stencil
of the diffusion problem dictates the structure of this graph and the states
to which a random walk may transition. Consider a random walk starting at node
$(i,j)$ in the mesh (the center node). If we play the Monte Carlo game to move
to a new state, then we may move to any of the nodes in this graph, including
the node at which we started with a probability proportional to the
coefficients of each node in Eq~(\ref{eq:fd_system}). If any node adjacent to
node $(i,j)$ is on another processor and the walk moves to that node, the
random walk must be communicated to that processor to continue the walk.

@@

We first compute the eigenvalues for the diffusion operator $\ve{D}$ by
applying the operator to the eigenfunctions and noting that $x=ih$ and $y=jh$
on a square Cartesian grid:
\begin{multline}
  \ve{D}\Phi_{p,q}(x,y) = \lambda_{p,q}(\ve{D})
  =\\ -\frac{D}{6h^2}\Big[4 e^{-2 \pi \imath p h} + 4 e^{2 \pi \imath
      p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2 \pi \imath q h} + e^{-2
      \pi \imath p h} e^{-2 \pi \imath q h} \\ + e^{-2 \pi \imath p h}
    e^{2 \pi \imath q h} + e^{2 \pi \imath p h} e^{-2 \pi \imath q h}
    + e^{2 \pi \imath p h} e^{2 \pi \imath q h} - 20\Big] + \Sigma_a
  \:.
  \label{eq:deriv_diff_1}
\end{multline}
Using Euler's formula, we can collapse the exponentials to
trigonometric functions:
\begin{equation}
  \lambda_{p,q}(\ve{D}) = -\frac{D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h) - 20] + \Sigma_a\:.
  \label{eq:deriv_diff_2}
\end{equation}

As Eq~(\ref{eq:diffusion_eq}) is diagonally dominant, Jacobi preconditioning
is sufficient to reduce the spectral radius of the iteration matrix below
unity and therefore ensure convergence of the Neumann series\footnote{Although
  $\rho(\ve{H}) < 1$ is a sufficient criteria for convergence of the Neumann
  series, recent work by Ji, Mascagni, and Li provides an additional necessary
  and sufficient condition for convergence of the Neumann-Ulam Monte Carlo
  algorithm \cite{ji_2013}. A diagonal scaling of the model problem also
  satisfies this condition.}. The preconditioner in this case is $\ve{M}
= diag(\ve{D})$ such that we are solving the following linear system:
\begin{equation}
  \ve{M}^{-1} \ve{D} \boldsymbol{\phi} = \ve{M}^{-1} \ve{s}\:.
  \label{eq:precond_diffsion}
\end{equation}
The operator $\ve{M}^{-1} \ve{D}$ is merely the original diffusion operator
with each row scaled by the diagonal component. As we have defined a
homogeneous domain, the scaling factor, $\alpha$, is the same for all rows in
the operator and defined as the $\phi_{i,j}$ coefficient from
Eq~(\ref{eq:fd_system}):
\begin{equation}
  \alpha = \Bigg[\frac{10 D}{3 h^2} + \Sigma_a\Bigg]^{-1}\:.
  \label{eq:jacobi_scaling}
\end{equation}
Using this coefficient, we have the following spectrum of preconditioned
eigenvalues:
\begin{equation}
  \lambda_{p,q}(\ve{M}^{-1} \ve{D}) = \alpha \lambda_{p,q}(\ve{D})\:.
  \label{eq:preconditioned_eigenvalues}
\end{equation}

@@

The adjoint Monte Carlo method is effectively an approximation to a
stationary method. Stationary methods for linear systems arise from
splitting the operator in Eq~(\ref{eq:linear_problem}) and iterating:
\begin{equation}
  \ve{x}^{k+1} = \ve{H}\ve{x}^k + \ve{c}\:,
  \label{eq:linear_iterative_method}
\end{equation}
with $k \in \mathbb{Z}^+$ defined as the \textit{iteration index} and $\ve{c}$
a modified source term. Defining $\ve{e}^k = \ve{u}^k - \ve{u}$ as the
solution error at the $k^{th}$ iterate, the error after $k$ iterations is:
\begin{equation}
  \ve{e}^{k} = \ve{H}^k\ve{e}^0\:. 
  \label{eq:linear_k_iter_error}
\end{equation}
By assuming $\ve{H}$ is diagonalizable \cite{leveque_finite_2007}, we have:
\begin{equation}
  ||\ve{e}^{k}||_2 \leq \rho(\ve{H})^k ||\ve{e}^0||_2\:.
  \label{eq:linear_k_iter_norm3}
\end{equation}

@@

In a domain decomposed algorithm, not all random walks will remain within the
domain they started in and must instead be communicated. This communication,
expected to be expensive, was analyzed by Siegel and colleagues for idealized,
load balanced situations for full nuclear reactor core Monte Carlo simulations
\cite{siegel_analysis_2012}.  To quantify the number of particles that leak
out of the local domain they define a leakage fraction, $\Lambda$, as:
\begin{equation}
  \Lambda = \frac{\text{average \# of particles leaving local domain}}
          {\text{total \# of particles starting in local domain}}\:.
          \label{eq:leakage_fraction}
\end{equation}
For their studies, it was assumed that the value of $\Lambda$ was dependent on
the total cross section of the system via the Wigner rational
approximation. Outlined more thoroughly by Hwang's chapter in
\cite{azmy_nuclear_2010}, we will use both the Wigner rational approximation
and the mean chord approximation as a means to estimate the leakage fraction.

In the case of parallel linear systems, we can use diffusion theory to
estimate the optical thickness of a domain in the decomposition and the
corresponding leakage fraction in terms of properties of the linear operator
and the discretization. We first calculate the mean distance a random walk
will move in the grid by computing the mean squared distance of its movement
along the chord of length $l$ defined along a cardinal direction of the domain
(i.e. $x$ or $y$). After a single transition a random walk will have moved a
mean squared distance of:
\begin{equation}
  \langle \bar{r_1^2} \rangle = (n_s h)^2\:,
  \label{eq:step_1_length}
\end{equation}
where $h$ is the size of the discrete grid elements along the chord and $n_s$
is the number of grid elements a random walk will move on average every
transition in the direction of the chord. For our diffusion model problem,
$n_s$ would equate to the expected number of states in the $i$ (or $j$ as the
problem is symmetric) direction that a random walk will move in a single
transition and is dependent on the stencil given in
Figure~\ref{fig:diffusion_graph}. After $k$ transitions, the random walk will
have moved a mean squared distance of:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k (n_s h)^2\:.
  \label{eq:step_k_length}
\end{equation}
If our chord is of length $l$ and there are $n_i$ grid elements (or
states to which a random walk may transition) along that chord, then $h =
l / n_i$ giving:
\begin{equation}
  \langle \bar{r_k^2} \rangle = k \Bigg(\frac{n_s l}{n_i}\Bigg)^2\:.
  \label{eq:step_k_length_sub}
\end{equation}
From diffusion theory, we expect the average number of interactions
along the chord to be:
\begin{equation}
  \tau = \frac{l}{2 d \sqrt{\langle \bar{r_k^2} \rangle}}\:,
  \label{eq:optical_thickness_1}
\end{equation}
where $d$ is the dimensionality of the problem and $\sqrt{\langle
  \bar{r_k^2} \rangle}$ is effectively the mean free path of the Monte
Carlo random walk in the domain. We can readily interpret $\tau$ to be the
\textit{effective optical thickness} of a domain of length
$l$. Inserting Eq~(\ref{eq:step_k_length_sub}) we get:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s \sqrt{k}}\:,
  \label{eq:optical_thickness_2}
\end{equation}
which if expanded with Eq~(\ref{eq:analytic_k}) gives us the final
relation for the effective optical thickness:
\begin{equation}
  \tau = \frac{n_i}{2 d n_s}
  \sqrt{\frac{\log(\rho(\ve{H}))}{\log(W_c)}}\:.
  \label{eq:optical_thickness_3}
\end{equation}

For optically thin domains, we expect that most random walks will be
communicated, while optically thick domains will leak the fraction of random
walks that did not interact within. Using the optical thickness defined in
Eq~(\ref{eq:optical_thickness_3}), we can complete the leakage approximations
by defining the bounds of $\tau \rightarrow 0, \Lambda \rightarrow 1$ and
$\tau \rightarrow \infty, \Lambda \rightarrow \tau^{-1}$.  With these bounds
we define the leakage fraction out of a domain using the Wigner rational
approximation:
\begin{equation}
  \Lambda = \frac{1}{1+\tau}\:,
  \label{eq:wigner_domain_leakage}
\end{equation}
and using the mean-chord approximation:
\begin{equation}
  \Lambda = \frac{1-e^{-\tau}}{\tau}\:.
  \label{eq:mean_chord_domain_leakage}
\end{equation}
Our derived leakage fraction is dependent on the eigenvalues of the iteration
matrix, the size of the domain, the content of the discretization stencil, and
the weight cutoff selected to terminate low weight random walks.

@@

In future work, these relationships will serve as guidelines for selecting an
appropriate parallel algorithm strategy and provide a basis for performance
models for future parallel implementations of the adjoint Neumann-Ulam
method. To be applicable to a broader range of problems, these models could
potentially be extended to non-symmetric and three-dimensional systems. In
addition, these relations will be used to analyze parallel implementations of
Monte Carlo synthetic acceleration methods that leverage the adjoint method
and serve as an initial grounds for assessing their feasibility for
large-scale problems.
